Isomorphisms as Groupoids: \\ Homotopy Type Theory and Fractional Types

### 2013-11-04 f.tex

@@

We propose to just consider functions are isomorphisms by construction. This
doesn't lose anything in terms of computability because we know that we could
do everything with reversible functions if we wanted to. There are actually
many advantages of doing so as we and others have argued at length
before. Anyway, the point is that if we have isomorphisms-by-construction
functions then the function itself becomes naturally a collection of
paths. We illustrate this point with examples.

@@

@@

@@

@@

>>>>

What can we say in pi about these two combinators:
\begin{verbatim}
Let f : A -> C and G : B -> D

(id x g) o (f x id)  :  A x B -> C x D

(f x id) o (id x g)  :  A x B -> C x D
\end{verbatim}
Are they the ``same''? We can reason extensionally to check that for all
inputs these two combinators produce the same results. In other words, we
reason from the axioms of category theory which includes things like
\verb|f o id = f| which are extensional equalities. (If we had weak
$n$-categories here what would happen, perhaps this is where we are leading
with the next sentence.) Better is to reason topologically exploiting the
original connections established by Joyal and
Street~\cite{planardiagrams,geometrytensor}. One can in this case show the
equivalence of these combinators by using the graphical language and noticing
that there is a \emph{planar isotopy} between the two diagrams, that moves
things around without crossing, cutting, or gluing any wires.

More formally, we view each type as a topological space and each function as
a deformation of the space expressed as a (sequential, parallel, etc.)
composition of primitive deformations (id, swap, unit, assoc, distribute,
factor, etc.). If we have a planar monoidal category, then there isn't much
you can do: the types are collections of lines grouped in bundles. The only
deformations that functions can do are to re-associate the bundles and get
rid of unit wires. (If the ambient space is a plane, i.e., if we restrict our
``hardware model'' to be a 2D model in which wires can't cross over each
other, then we can't even equate the functions that apply id to different
bundles. If the ambient space is 3D then we can. Eq. 3.2 in Selinger's survey
of graphical languages for monoidal categories.) An example of two functions
we can show to be the same in the planar model is the one above in the
verbatim block. An isotopy exists between the two combinators because we can
move things around in the plane as long as we don't cross or cut any wires.

If we allow braiding (i.e, commutativity) we must move to ambient isotopies
in 3D. In particular, the two versions of not that cross true over false or
vice versa are not isotopic.

If we also want, the composition of not and not to be expressible directly as
a twist, we must move to \emph{framed} isotopies in 3 dimensions, where the
lines are represented as flat ribbons.

A symmetric monoidal category requires ambient isotopy in 4 dimensions so
we're not going to consider it.

If we are going to have higher-order functions, we need to move up, i.e., we
need to build a space in which the types are the functions of the lower
level, and in which the functions are the isotopies of the lower level. In
other words, we need to be able to express the lower level functions as
topological spaces and we need to be able to express the lower level
isotopies as functions.

Think of a function as a map connecting points from one space to points in
the other space (respecting paths etc.). This map gives a geometric shape. We
will think of one line in that shape as solid with all the other lines dashed.

Another unrelated observation is that Pi gives us an inductive
characterization of functions! If that whole thing with fractionals
works out we will have an inductive characterization of higher-order
functions!

@@

We propose to just consider functions that are isomorphisms by
construction. This doesn't lose anything in terms of computability because we
know that we could do everything with reversible functions if we wanted
to. There are actually many advantages of doing so as we and others have
argued at length before. Anyway, the point is that if we have
isomorphisms-by-construction functions then the function itself becomes
naturally a collection of paths. We illustrate this point with examples.

@@

Are higher-order inductive types something like ADTs where we have equations
or something like graphs where we can build interesting shapes inductively or
perhaps both somehow?

@@

Double negation vs. braids and twists. Any insights from category theory or
topology that can be used to understand classical proofs?

@@

Connections between standard circuit model and topological quantum computing.

@@

CS: Give computational content of science
  Logic: Curry-Howard
  Classical logic: Continuations
  Physics: Quantum information; computation
  cognition
  linguistics
  Construction of cellos

Mathematical proofs have computational content
Physical equations have computational content

Now at the heart of it all: EQUALITY has computational content and is the
most fundamental relation we use implicitly

Theory at the center of many many things: topology, algebra, geometry, type
theory, physics, etc. ==> Exciting

Conservation of information, energy, etc. 
Symmetry
F = m a
or E = m C^2
are not EQUALITIES. They are isomorphisms: go from one to the other
preserving various entitites. 
Get rid of equality ==> use isomorphisms exclusively

Turtles all the way down (or up in that case)

when are two structures "the same"
if their elements are the same

what is the structures are infinite? (i.e., described by a process); go
right, then right, then right, then right is the same as stay here but not
quite the same

E=mc2, F=ma

Leibnitz

 - topology of computation/evaluation: in addition to loops/circle,
   bernstein conditions for parallelism; or more accurately two
   computations are homotopic if one is sequential and is parallel and I
   can massage one to the other

 - univalence: (A \== B) isEquiv (A isEquiv B) fun. ext. (f \== g) -> (x
   : A) f(x) \== g(x) is an equivalence paths between complex types iff
   there are paths between the points: works for simple finite types;
   for infinite things can't be proved; need these two additional axioms

 - for my intro: what is hott; article about hott in magazine; 
   steve awodey's talk
   "http://www.andrew.cmu.edu/user/awodey/hott/CMUslides.pdf"
   formalize mathematics if you care about that: eckmann-hilton
   "http://blogs.scientificamerican.com/guest-blog/2013/10/01/voevodskys-mathematical-revolution/"
    Logic level -1
    Set theory level 0
    groupoid level 1
    ...
    The Propositions-as-Types conception of Martin-Löf type theory leads
   to a system of logic that is different from both classical and
   intuitionistic logic with respect to the statements that hold in the
   fragment of logical language that these three share, namely the
   quantifiers \Pi , \Sigma , equality \mathsf{Id} , conjunction \times
   , disjunction + , implication \rightarrow and negation \neg .  For
   example, it validates the axiom of choice but not the law
   of excluded middle  a combination that is impossible
   classically and intuitionistically.  Let us call this conception
   constructive validity following a paper by Dana Scott in
   which it was first introduced (Constructive Validity, Symposium on
   Automated Deduction, LNM 125, Springer (1970), pp. 237-275).

 - for class "http://math.ucr.edu/home/baez/eckmann_hilton.jpg"

 - circle ==> recursion: homotopy between path p and path q means
   if we can move from p to q without getting snagged in a loop, i.e.,
   without crossing over an infinite loop!!

 - Agda and homotopy type theory idea: Ulf showed us how to fill in
   "small" program fragments automatically. If we had more precise types
   can the compiler fill in different algorithms for us. Here is an
   idea: in HoTT we can control the isomorphisms and what we mean by
   them. For example, we can add paths between A and B to mean that B is
   a sorted version of A. Then different arrows would correspond to
   different sorting algorithms. If we maintain proof relevance then we
   can do actions depending on the particular sorting algorithm that is
   used to go from A to B. This might be doable in Pi actually.

- different levels correspond to fractal equality or perhaps each level gives
  you more precision

- Don't you think that these:
  "http://en.wikipedia.org/wiki/Loop_optimization"
  are homotopies. They do make some explicit connections to
  "geometry/topology". I think we have discovered a gold mine

So I thought more about it and I think it holds more water. Here is the story:

- paths in HoTT represent computations; this is true in the conventional
  HoTT but even more or so in our context where all computations are
  isomorphisms;

- paths that loop are therefore looping computations;

- the loop in a 2groupoid terminates after two iterations;

- the loop in S1 is an infinite loop;

- the loop in S2 is a double infinite loop (on two natural
  numbers);

- the various geometric objects capture various patterns of iteration!

- Filinski's paper below shows that if you are given one simple infinite
  loop and continuations, then you can expression recursion.  So if we
  include S1 in our system and we figure out how to do higher-order
  functions (and their duals as delimited continuations), then we should
  be able to express recursion. A small hope would be that we can get
  away without full continuations and that we can get recursion from S1
  using just the groupoid structure but that's speculative.

I like it! --Amr

Check this out: "http://iml.univ-mrs.fr/geocal06/"=20

In particular this workshop:

Algebraic Topology, Concurrency and Rewriting

Organizers: =C3=89ric Goubault, Yves Lafont

The objective of the course is to give a general overview of the recent
developments of ``geometrical'' approaches to the theory of computation,
in particular concurrent and distributed computations, and rewriting.

The course will start by a general introduction to homotopy and
homology, classical on topological spaces, and generalize through
abstract homotopy theory and generalized homology theories. Then two
different applications will be developped.

The first one concerns concurrency theory. In this application, a
refined form of homotopy (dihomotopy) has to be designed, that takes
into account the direction of time. Similar methods as the ones on
ordinary topological spaces can be defined, in order to find
``invariants'' under dihomotopy, i.e. a way to classify directed shapes
under (directed) deformation. These invariants can in turn be used for
purposes of proofs or static analysis of algorithms or programs. We will
show such applications in the course. Other applications, of more
classical algebraic topological invariants will also be derived, in the
field of fault-tolerant distributed systems.

The second one concerns the rewriting theory. Since the pioneering work
by Squier, we know how to use homology to characterize the algebraic
objects presented by rewriting systems. Basically, words are seen as
objects of dimension one, rewriting rules are seen as of dimension two,
confluence diagrams as of dimension three etc. providing resolutions to
the algebraic objects they present. Extensions to term rewriting has
been recently solved, along similar lines, using more general homology
theories. Also higher-dimensional rewriting has gained wide interest and
its theory will be fully developped, in conjunction with the necessary
n-categorical (or polygraphic) theory.

mfsp paper by Fajstrup, Goubault, and Raussen

http://www.pps.univ-paris-diderot.fr/~laurent/geocal/slides/mellies.pdf‎

-- Repeat for monoids without using equivalences!

-- Let's build a monoidal category whose objects are ∞-groupoids using
-- Thorsten's trick perhaps. Or perhaps we only need 1-groupoids or
-- 2-groupoids... Or better use HITs: explicit constructors for points
-- and paths and let the whole ∞-groupoid structure be implicit and
-- generated by J on demand as needed. A morphim between such HITs
-- would map points to points and paths to paths LYING OVER the
-- original paths (using transport).

-- multiplicative monoid:
-- bifunctor ⊗ maps two types A and B to the type (A ⊗ B):
--   map a point in A and a point in B to a point in (A ⊗ B)
--   map a path in A and a path in B to a path in (A ⊗ B)
--   if paths p in A and q in B are respectively equal to paths p' in
--     A and q' in B then their images in (A ⊗ B) are also equal
-- associative up to ...
-- an object I which is both a left and right identity for ⊗ 
-- coherence conditions: pentagon and triangle
-- 
-- Now what is the structure on 1 and ⊗ in terms of paths on the
-- components

-- map points to points
-- map paths to paths
-- preserve ≡

apTensor : {A B C : Set} {x y : A × B} → (f : A × B → C) → (x ≡ y) → (f x ≡ f y)
apTensor {A} {B} {C} {x} {y} f p = 
  pathInd 
    (λ {x} {y} p → f x ≡ f y) 
    (λ x → refl (f x))
    {x} {y} p

-- monoid 0 and ⊕

-- do all these sections in Ch. 2 with monoidal categories instead of
-- the usual type constructors

-- imagine we had tensor products instead would be it possible to
-- convert a path : A × B → C × D to two sequential paths A → C with
-- refl on B followed by the path B → D and refl on C. So the
-- topological view of computation would express which dependencies
-- allow you to move computations around to make them more or less
-- parallel. Delooping sounds like it would allow you to make a loop
-- (sequential) into something that operates in parallel (say every
-- two iterations happen in parallel) but that can only happen if the
-- dependencies are "right".

loops/spaces; loop optimizations; cata/epi/etc-morphisms are also
patterns of recursion; probably even parametricity and logical
relations; all these things must be homotopies.

1/x gives you the pattern to define a clause
+ gives you the ability to write all the clauses
- gives you the ability to retract a clause
all functions are constructible from the identity by massaging it
functors between types internalized inside a type using 1/., +, and -

PI

cardinality of these infinity groupoids

delooping: 1/Bool = 1/(delooping false) + 1/(deloopoing true)

≡ {u+1} {Set u : Set (u+1)} (T1 : Set u) (T2 : Set u) -> Set (u+1)

CC : ∀ {u} → (A B : Set u) → _≡_ {L.suc u} {Set u} A B → Set (L.suc u)
CC {u} A B p = _≡_ {L.suc u} {Set u} (A × B) (B × A)

cc : ∀ {u} → (A : Set u) → _≡_ {L.suc u} {Set u} (A × A) (A × A)
cc A = refl

swapP : ∀ {u} → (A B : Set u) → _≡_ {L.suc u} {Set u} (A × B) (B × A) 
swapP {u} = pathIndU {u} {Set u} CC cc

t : ℕ × Bool ≡ Bool × ℕ
t = swapP ℕ Bool

How many elements are in this set:

{ Egyptian, American, Chinese }

It depends on the equivalence relation. If you think the trivial
relation, there are 3 elements. If you think the relation that
identifies Egyptian/American (Amr) then you have two elements only in
the set. We can make this equivalence relation depend on the
particular element in question and we can make it first-class. When
viewed by itself, it is like the fraction 2/3: it takes 3 elements and
identifies two of them giving you back two elements.

https://inf.ug.edu.pl/~stefan/AlgTop/

All programming only builds trees. 
Graphs require assignments and ugly things
HoTT allows us to build datatypes with "paths", "shapes", etc.

CS -> Topology: study shapes with inductive types
Topology -> CS: richer datatypes with cycles

Can I build types that solve polynomials ???

Define pointed types (A,a)

Reciprocal of a pointed type (A,a) is delooping (a=a,refl)

delooping of A is (B,b) such that loopspace(B,a) = A

our construction of fractionals is delooping: take n points connected by a
group stucture; turn them into a point with the group structure turned into
path composition.
